---
title: "Organize Data Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
# loading the raw dataset
df_raw_wdi <- read_csv(here("data", "data_raw", "wdi_data_original.csv"))

# displaying the first few rows to inspect the data
head(df_raw_wdi)

# displaying a summary to get an overview of the variables
summary(df_raw_wdi)

```
# Raw data to ---------> analytical data
# check yourself. before wreck yourself
```{r}
# --- 2. Data Cleaning and Transformation ---

# 2.1. Here, renaming columns to use meaningful and consistent names.
# mapping the CSV column names to the variable names I want.
# Note: Both 'cpi' and 'inflation' are in my CSV. using 'inflation' as the annual % change,
# since it's more appropriate as a macroeconomic stability indicator. The CPI index is just for reference.
cleaned_data <- df_raw_wdi |>
  janitor::clean_names() |>
  dplyr::rename(
    eodb_score = edb, # ease of doing Buisness indicator
    access_electricity = electricity,
    broadband_internet_users_pct = broadband, # renaming 'broadband' for clarity
    mobile_subscriptions = mobile,
    gdp_per_capita = gdp_pc,
    current_account_balance = cab,
    inflation_annual_pct = inflation, # using 'inflation' as annual % change
    cpi_index = cpi, # keeping CPI index just for reference
    interest_rate_spread = interest_spread,
    government_effectiveness = gov_effectiveness,
    rule_of_law_estimate = rule_law,
    control_of_corruption_estimate = control_corruption
    
    # and sticking to a minimal set of non-overlapping variables
  )

# 2.2. dropping rows where my dependent variable (eodb_score) is NA,
# and filtering out aggregate income rows to keep only actual countries.
cleaned_data <- cleaned_data |>
  drop_na(eodb_score) |>
  filter(income != "Aggregates")

# 2.3. Here, selecting only the final set of variables for my analysis.
# excluding metadata like 'status' and 'lastupdated',
# and any geographical metadata that I don't need for the core analysis, but keeping country identifiers.
analytical_dataset <- cleaned_data|>
  select(-status,-iso2c, -lastupdated, -lending, -longitude, -latitude,-interest_rate_spread)

```

# visualaize 

```{r}
ggplot(analytical_dataset, aes(x = income, y = eodb_score)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6) +
  labs(title = "Ease of Doing Business (EDB) Score by Income Group",
       x = "Income Group", y = "EDB Score") +
  theme_minimal()

```

```{r}
ggplot(analytical_dataset, aes(x = income, y = eodb_score)) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "darkblue") +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  labs(title = "EDB Score by Income Group",
       x = "Income Category", y = "EDB Score") +
  theme_minimal()

```



# descriptive table


```{r}
datasummary_skim(analytical_dataset)
```



# Tidy data
# Key dvs- # narrow Down to One dependent variable
# Key IVS
#few important Covariables
# 
```{r}
library(corrplot)

# Create correlation matrix with piping, then plot it
analytical_dataset |>
  select(-year) |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs") |>
 corrplot(method = "color", 
         type = "upper", 
         order = "hclust", 
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.7,
         title = "Correlation Heatmap of Business Environment Factors",
         addCoef.col = "black",   
         number.digits = 1,       
         number.cex = 0.5)    
```

# data wrangling

#missing values

# i am not imputing any data
# dropping the missing ones- As i dont have that much of missing data variables.

# merging data

# reshaping data (not in may case)



```{r}

save(analytical_dataset, file = here("data", "data_constructed", "analytical_data.RData"))

```

